Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-17 output/latex/manuscript.tex
use "Supplementary Figure/Table" and reset figure/table counter in supplementary information section

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms},
  pdfauthor={Milton PividoriSumei LuBinglan LiChun SuMatthew E. JohnsonWei-Qi WeiQiping FengBahram NamjouKrzysztof KirylukIftikhar KulloYuan LuoBlair D. SullivanBenjamin F. VoightCarsten SkarkeMarylyn D. RitchieStruan F.A. GrantCasey S. Greene},
  pdflang={en-US},
  pdfkeywords={genetic studies, functional genomics, gene co-expression, therapeutic targets, drug repurposing, clustering of complex traits},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

@@

Finally, we ran our GLS model (Equation (\ref{eq:reg:model})) to compute an association between each of the 987 LVs in MultiPLIER and the 1,000 S-MultiXcan results on random phenotypes.
For this, we built a gene correlation matrix specifically for this cohort (see \protect\hyperlink{sec:methods:reg}{Methods}).
Then, we compared the GLS results with an equivalent, baseline ordinarly least squares (OLS) model assuming independence between genes.
Figure \ref{fig:reg:nulls:qqplots} compares the distribution of \(p\)-values of the OLS and GLS models.
The GLS model has a slightly smaller mean type I error rate (0.0558, SD=0.0127) than the baseline OLS model (0.0584, SD=0.0140), and \(p\)-values follow more closely the expected uniform distribution.
Importantly, the GLS model is able to correct for LVs with adjacent and highly correlated genes at the top such as LV234 (Figure \ref{fig:reg:nulls:qqplot:lv234}), LV847 (Figure \ref{fig:reg:nulls:qqplot:lv847}), LV45 (Figure \ref{fig:reg:nulls:qqplot:lv45}), or LV800 (Figure \ref{fig:reg:nulls:qqplot:lv800}), among others.
In contrast and as expected, the OLS model has higher mean type I errors and smaller-than-expected \(p\)-values in all these cases.

We also detected other LVs with higher-than-expected mean type I errors for both the GLS and OLS models, although they don't have a relatively large number of adjacent genes at the top.
One example is LV914, shown in Figure \ref{fig:reg:nulls:qqplot:lv914}.
Inflation in these LVs might be explained by inaccuracies in correlation estimates between the individual-level MultiXcan model and its summary-based version (see Methods).
Therefore, we flagged those with a type I error rate larger than 0.07 (127 LVs) and excluded them from our main analyses.
We didn't see signs of inflation when applying the method in real data (Figure \ref{fig:reg:real:qqplots}).

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

The results of this analysis (Figure \ref{fig:sup:clustering:agreement}) show that, under the two simulated null scenarios, the agreement of the consensus partitions with the ensemble is very close to zero.
This means, as expected, that there is no consensus among ensemble partitions generated with different clustering algorithms and data representations.
In contrast, using the real data, the consensus clustering approach finds trait pairs that are grouped together across the different members of the ensemble.
The partitions above the 75th percentile were considered in the main analyses, and are shown in the clustering tree in Figure \ref{fig:clustering:tree}.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

%% pandoc-fignos: environment for tagged figures
\newenvironment{fignos:tagged-figure}[1][]{
  \let\oldfigurename\figurename
  \renewcommand{\figurename}{Supplementary Figure}
}{
  \let\figurename\oldfigurename
}

%% pandoc-tablenos: environment for tagged tables
\newenvironment{tablenos:tagged-table}[1][]{
  \let\oldtablename\tablename
  \renewcommand{\tablename}{Supplementary Table}
}{
  \let\tablename\oldtablename
}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms},
  pdfauthor={Milton PividoriSumei LuBinglan LiChun SuMatthew E. JohnsonWei-Qi WeiQiping FengBahram NamjouKrzysztof KirylukIftikhar KulloYuan LuoBlair D. SullivanBenjamin F. VoightCarsten SkarkeMarylyn D. RitchieStruan F.A. GrantCasey S. Greene},
  pdflang={en-US},
  pdfkeywords={genetic studies, functional genomics, gene co-expression, therapeutic targets, drug repurposing, clustering of complex traits},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\setcounter{figure}{0}

@@

Finally, we ran our GLS model (Equation (\ref{eq:reg:model})) to compute an association between each of the 987 LVs in MultiPLIER and the 1,000 S-MultiXcan results on random phenotypes.
For this, we built a gene correlation matrix specifically for this cohort (see \protect\hyperlink{sec:methods:reg}{Methods}).
Then, we compared the GLS results with an equivalent, baseline ordinarly least squares (OLS) model assuming independence between genes.
Supplementary Figure \ref{fig:reg:nulls:qqplots} compares the distribution of \(p\)-values of the OLS and GLS models.
The GLS model has a slightly smaller mean type I error rate (0.0558, SD=0.0127) than the baseline OLS model (0.0584, SD=0.0140), and \(p\)-values follow more closely the expected uniform distribution.
Importantly, the GLS model is able to correct for LVs with adjacent and highly correlated genes at the top such as LV234 (Supplementary Figure \ref{fig:reg:nulls:qqplot:lv234}), LV847 (Supplementary Figure \ref{fig:reg:nulls:qqplot:lv847}), LV45 (Supplementary Figure \ref{fig:reg:nulls:qqplot:lv45}), or LV800 (Supplementary Figure \ref{fig:reg:nulls:qqplot:lv800}), among others.
In contrast and as expected, the OLS model has higher mean type I errors and smaller-than-expected \(p\)-values in all these cases.

We also detected other LVs with higher-than-expected mean type I errors for both the GLS and OLS models, although they don't have a relatively large number of adjacent genes at the top.
One example is LV914, shown in Supplementary Figure \ref{fig:reg:nulls:qqplot:lv914}.
Inflation in these LVs might be explained by inaccuracies in correlation estimates between the individual-level MultiXcan model and its summary-based version (see Methods).
Therefore, we flagged those with a type I error rate larger than 0.07 (127 LVs) and excluded them from our main analyses.
We didn't see signs of inflation when applying the method in real data (Supplementary Figure \ref{fig:reg:real:qqplots}).

\begin{fignos:tagged-figure}[S1]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S2]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S3]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S4]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S5]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S6]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S7]

@@

\end{fignos:tagged-figure}

@@

\begin{fignos:tagged-figure}[S8]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S9]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S10]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S11]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S12]

@@

\begin{tablenos:tagged-table}[S1]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S2]

@@

\end{tablenos:tagged-table}

@@

\begin{fignos:tagged-figure}[S13]

@@

\end{fignos:tagged-figure}

The results of this analysis (Supplementary Figure \ref{fig:sup:clustering:agreement}) show that, under the two simulated null scenarios, the agreement of the consensus partitions with the ensemble is very close to zero.
This means, as expected, that there is no consensus among ensemble partitions generated with different clustering algorithms and data representations.
In contrast, using the real data, the consensus clustering approach finds trait pairs that are grouped together across the different members of the ensemble.
The partitions above the 75th percentile were considered in the main analyses, and are shown in the clustering tree in Figure \ref{fig:clustering:tree}.

@@

\begin{fignos:tagged-figure}[S14]

@@

\begin{tablenos:tagged-table}[S3]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S4]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S5]

@@

\begin{tablenos:tagged-table}[S6]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S7]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S8]

@@

\begin{tablenos:tagged-table}[S9]

@@

\begin{tablenos:tagged-table}[S10]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S15]

@@

\begin{tablenos:tagged-table}[S11]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S16]

@@

\begin{tablenos:tagged-table}[S12]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S13]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S14]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S17]

@@

\begin{tablenos:tagged-table}[S15]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S16]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S17]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S18]

@@

\begin{tablenos:tagged-table}[S18]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S19]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S20]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S19]

@@

\begin{tablenos:tagged-table}[S21]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S22]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S23]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S20]

@@

\begin{tablenos:tagged-table}[S24]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S25]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S26]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S21]

@@

\begin{tablenos:tagged-table}[S27]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S28]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S29]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S22]

@@

\begin{tablenos:tagged-table}[S30]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S31]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S32]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S23]

@@

\begin{tablenos:tagged-table}[S33]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S34]

@@

\end{tablenos:tagged-table}

\begin{tablenos:tagged-table}[S35]

@@

\end{tablenos:tagged-table}

\begin{fignos:tagged-figure}[S24]
